---
title: "analysis"
output: html_document
date: '2022-09-14'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyr)
library(MASS)
library(lubridate)
library(rjags)
library(INLA)
```
Interesting questions: competing risks: a lot of people died during covid; at some point we should see a decline in overall death rates below baseline because of this 'harvesting' effect. in which groups will this be most apparent?
https://www.bmj.com/content/373/bmj.n1137

Kissler and Grad: https://academic.oup.com/aje/article/191/8/1519/6572389

Analysis idea; define excess death rate by 5 year age bin. (e.g., simple approach, annual incidence in previous year vs 2020 with trend adjustment; annual deaths; use bayesian model)

Deaths in year y, age group i:
 N_deaths_y_i ~ Poisson (lambda_y_i)
 log(lambda_y_i) = log(pop_y_i) + alpha_i +beta1_i*t + beta2_i*ind2020
 beta1_i ~dnorm(rho1*beta1_(i-1), tau) 
 
```{r}
a1 <- readRDS('./Data/all_cause_deaths_age_race_sex_year.rds')
#Uses age recode 27
#https://www.cdc.gov/nchs/data/dvs/Multiple-Cause-Record-Layout-2020.pdf 

a2 <- a1 %>%
  group_by(agey,race_recode, month, year,sex) %>%
  mutate(agey = if_else(agey>85,85, agey)) %>%
  filter(agey !=999) %>%
  summarize(N_deaths=sum(N_deaths)) %>%
  ungroup() %>%
    tidyr::complete(agey,race_recode, month, year,sex, fill=list(N_deaths=0)) %>% #fills 0s
  mutate(date=as.Date(paste(year, month, '01', sep='-'), format='%Y-%m-%d'),
    agec = if_else(agey<35,1,
           if_else( agey<45,2,
           if_else( agey<55,3,
           if_else( agey<65,4,
           if_else( agey<75,5,
           if_else( agey<85,6,7                          ))))))
  )


pop1 <- readRDS('./Data/pop_interpol_single_age.rds')

a3 <- full_join(a2, pop1, by=c('agey','sex','race_recode','date')) %>%
  mutate(    agec = as.factor(agec),
    race_recode= as.factor(race_recode)
    ) %>%
  filter( !is.na(pop.interpol) & !is.na(pop.interpol) & year >= 2014) %>%
  mutate(N_deaths = if_else(is.na(N_deaths),0, N_deaths),
         month=as.factor(month(date)),
         pandemic_period= if_else(date>='2020-03-01',1,0),
         year= year(date))
```

  
   RACE_RECODE: 
    1=Non-Hispanic White
   2=Non-Hispanic- Black
   3= Hispanic
   4=American Indian/Native Alaskan
   5= Asian/Pacific Island
   999=Missing
```{r}

mod1 <- glm.nb(N_deaths ~ pandemic_period*race_recode + 
              pandemic_period*sex + 
              year + 
              month + 
              pandemic_period*agec + 
              offset(log(pop.interpol)),  data=a3  )

summary(mod1)


```

```{r}

mod.mat <- model.matrix(~ pandemic_period + race_recode + agey +sex, data=a3)

model_string <- "
model{
for(t in 1:N_times){ 
  N_deaths[t] ~ dpois(lambda[t])
  
  log(lambda[t]) <- ( int )
  
  }
   int ~ dnorm(0, 1e-4)
   
 }
"
##############################################################
#Model Fitting
##############################################################
inits1=list(".RNG.seed"=c(123), ".RNG.name"='base::Wichmann-Hill')
inits2=list(".RNG.seed"=c(456), ".RNG.name"='base::Wichmann-Hill')
inits3=list(".RNG.seed"=c(789), ".RNG.name"='base::Wichmann-Hill')
##############################################
#Model Organization
##############################################
model_spec<-textConnection(model_string)
model_jags<-jags.model(model_spec, 
                       inits=list(inits1,inits2, inits3),
                       data=list('N_deaths'=a3$N_deaths ,
                                 'N_times'=nrow(a3) 
                                    ),
                       n.adapt=10000, 
                       n.chains=3)
params<-c('lambda', 'beta','epsilon', 'rho3','tau3.beta', 'int')
##############################################
#Posterior Sampling
##############################################
posterior_samples<-coda.samples(model_jags, 
                                params, 
                                n.iter=10000)
posterior_samples.all<-do.call(rbind,posterior_samples)
#post1.summary<-summary(posterior_samples)
#post_means<-colMeans(posterior_samples.all)
post_means<-apply(posterior_samples.all, 2, mean)
sample.labs<-names(post_means)
ci<-t(hdi(posterior_samples.all, credMass = 0.95))
#ci<-matrix(sprintf("%.1f",round(ci,1)), ncol=2)
ci<-matrix(ci, ncol=2)
row.names(ci)<-sample.labs
#post_means<-sprintf("%.1f",round(post_means,1))
names(post_means)<-sample.labs
indices <- str_extract_all(names(post_means), "(?<=\\[).+?(?=\\])")
rep1 <- sapply(indices, function(x) identical(x, character(0))) #is it missing ie character(0)
indices[which(rep1==1)] <- '999'

```


 
 