---
title: "analysis"
output: html_document
date: '2022-09-14'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyr)
library(MASS)
library(lubridate)

library(plotly)
library(INLA)
library(dplyr, warn.conflicts = FALSE)

# Suppress summarise info
options(dplyr.summarise.inform = FALSE)

library(rjags)
library(stringr)
library(ggplot2)
library(HDInterval)
library(readxl)
library(patchwork)
library(INLA)
source('./R/utils_kissler.R')
```
Interesting questions: competing risks: a lot of people died during covid; at some point we should see a decline in overall death rates below baseline because of this 'harvesting' effect. in which groups will this be most apparent?
https://www.bmj.com/content/373/bmj.n1137

Kissler and Grad: https://academic.oup.com/aje/article/191/8/1519/6572389

Analysis idea; define excess death rate by 5 year age bin. (e.g., simple approach, annual incidence in previous year vs 2020 with trend adjustment; annual deaths; use bayesian model)

Deaths in year y, age group i:
 N_deaths_y_i ~ Poisson (lambda_y_i)
 log(lambda_y_i) = log(pop_y_i) + alpha_i +beta1_i*t + beta2_i*ind2020
 beta1_i ~dnorm(rho1*beta1_(i-1), tau) 
 
```{r}
a1 <- readRDS('./Data/all_cause_deaths_age_race_sex_year.rds')
#Uses age recode 27
#https://www.cdc.gov/nchs/data/dvs/Multiple-Cause-Record-Layout-2020.pdf 

a2 <- a1 %>%
  group_by(agey, month, year,sex) %>%
  mutate(agey = if_else(agey>120,120, agey)) %>%
  filter(agey !=999) %>%
  summarize(N_deaths=sum(N_deaths)) %>%
  ungroup() %>%
    tidyr::complete(agey, month, year,sex, fill=list(N_deaths=0)) %>% #fills 0s
  mutate(date=as.Date(paste(year, month, '01', sep='-'), format='%Y-%m-%d'),
    agec = if_else(agey<35,1,
           if_else( agey<45,2,
           if_else( agey<55,3,
           if_else( agey<65,4,
           if_else( agey<75,5,
           if_else( agey<85,6,7                          ))))))
  )


#pop1 <- readRDS('./Data/pop_interpol_single_age.rds')
#https://www.census.gov/newsroom/press-kits/2020/population-estimates-detailed.html
#https://www.census.gov/data/tables/time-series/demo/popest/2020s-national-detail.html

pop1 <- read_excel('./Data/nc-est2019-syasexn.xlsx', sheet='compiled') %>%
  mutate(age=round(age*100)) %>%
 reshape2::melt( id.vars=c('age', 'sex')) %>%
  mutate(year=as.numeric(substring(variable,2)), month=7)

  months <- cbind.data.frame('month'=1:12)

  pop1a <-    bind_rows(pop1, months) %>%
    tidyr::complete( month, nesting(year,age,  sex), fill=list(pop=NA) ) %>%
    mutate(date= as.Date(paste(year, month, '01', sep='-'))) %>%
    filter(date>='2013-07-01')  %>%
    rename(pop=value, agey=age) %>%
    arrange(agey , sex,year, month) %>%
    group_by(agey, sex) %>%
    mutate(time= year + month/12 - 1/12 ) %>%
    mutate(pop.interpol=approx(time,pop,time)$y) %>%
    ungroup()%>%
    select(agey, sex, date, pop.interpol)

a3 <- full_join(a2, pop1a, by=c('agey','sex','date')) %>%
  mutate(    agec = as.factor(agec)
    ) %>%
  filter( !is.na(pop.interpol) & year >= 2014) %>%
  mutate(N_deaths = if_else(is.na(N_deaths),0, N_deaths),
         month=as.factor(month(date)),
         pandemic_period= if_else(date>='2020-04-01',1,0), #START APRIl 1 (QTR2)
         year= year(date))
```

  
   RACE_RECODE: 
    1=Non-Hispanic White
   2=Non-Hispanic- Black
   3= Hispanic
   4=American Indian/Native Alaskan
   5= Asian/Pacific Island
   999=Missing
   
```{r}

# mod1 <- glm.nb(N_deaths ~ pandemic_period*race_recode + 
#               pandemic_period*sex + 
#               year + 
#               month + 
#               pandemic_period*agec + 
#               offset(log(pop.interpol)),  data=a3  )
# 
# summary(mod1)


```


## Single group by age
```{r}


b1 <- a3 %>%
  mutate(qtr=as.factor(quarter(date))) %>%
  group_by(year,month,qtr,pandemic_period, agey) %>%
    summarize(N_deaths=sum(N_deaths), pop=sum(pop.interpol)) %>% 
  ungroup() %>%
  arrange(agey, year, qtr) %>%
  group_by(agey,year,qtr,pandemic_period) %>%
  summarize(N_deaths=sum(N_deaths), pop=mean(pop)) %>% # for the quarters, take mean so don't double count people
  mutate(log.offset=log(pop/100000), 
         raw_prob_death= N_deaths/pop,) %>% 
  ungroup() %>%
  group_by(agey) %>%
  mutate(t=row_number(), agey2=agey, agey3=agey) %>%
  ungroup() %>%
  arrange(year, qtr,pandemic_period, agey )



# b1 <- a3 %>%
#   mutate(qtr=as.factor(quarter(date))) %>%
#   group_by(year,month,qtr,pandemic_period, agey) %>%
#     summarize(N_deaths=sum(N_deaths), pop=sum(pop.interpol)) %>% 
#   ungroup() %>%
#   group_by(year,qtr,pandemic_period, agey) %>%
#   summarize(N_deaths=sum(N_deaths), pop=mean(pop)) %>% # for the quarters, take mean so don't double count people
#   mutate(log.offset=log(pop/100000), raw_prob_death= N_deaths/pop)




```

  

```{r}
#run the model
source('./R/inla_mod.R')
```

```{r}
# Extract fitted values
b1$pred1 <- inla.out$summary.fitted.values[,"mean"]

p0.fitted <- ggplot(b1, aes(x=agey, y=log(pred1/pop), group=t, col=t))+
  geom_line()+
  theme_classic()

#extract random effects
re.age <- inla.out$summary.random$agey %>%
  mutate()

p1.age.inla <- ggplot(re.age, aes(x=ID, y=mean))+
  geom_line()+
  theme_classic()+
  geom_ribbon(data=re.age, aes(x=ID, ymin=`0.025quant`, ymax=`0.975quant`), alpha=0.2)
p1.age.inla

re.age.time <- inla.out$summary.random$agey2 %>%
  mutate()

p2.time.age.inla <- ggplot(re.age.time, aes(x=ID, y=exp(mean)))+
  geom_line()+
  theme_classic()+
  geom_ribbon(data=re.age.time, aes(x=ID, ymin=exp(`0.025quant`), ymax=exp(`0.975quant`)), alpha=0.2)+
  geom_hline(yintercept = 1)
p2.time.age.inla

re.age.pandemic <- inla.out$summary.random$agey3 %>%
  mutate()

p3.pandemic.age.inla <- ggplot(re.age.pandemic, aes(x=ID, y=exp(mean)))+
  geom_line()+
  theme_classic()+
  geom_ribbon(data=re.age.pandemic, aes(x=ID, ymin=exp(`0.025quant`), ymax=exp(`0.975quant`)), alpha=0.2) +
  geom_hline(yintercept = 1)
p3.pandemic.age.inla


# Fixed effects
summary(inla.out)$fixed

# Random effects
summary(inla.out)$hyperpar
```


```{r, eval=F}
#RUN THE JAGS MODEL
#source('./R/jags_mod.R')
```


trouble with epsi lon an dbeta corr
```{r}
posterior_samples.all <- readRDS( './Results/posteriors.rds')

#post1.summary<-summary(posterior_samples)
#post_means<-colMeans(posterior_samples.all)

post_means<-apply(posterior_samples.all, 2, mean)
sample.labs<-names(post_means)
ci<-t(hdi(posterior_samples.all, credMass = 0.95))
#ci<-matrix(sprintf("%.1f",round(ci,1)), ncol=2)
ci<-matrix(ci, ncol=2)
row.names(ci)<-sample.labs
#post_means<-sprintf("%.1f",round(post_means,1))
names(post_means)<-sample.labs

indices <- str_extract_all(names(post_means), "(?<=\\[).+?(?=\\])")
rep1 <- sapply(indices, function(x) identical(x, character(0))) #is it missing ie character(0)
indices[which(rep1==1)] <- '999'
index_n <- as.numeric(unlist(indices))

combined <- cbind.data.frame(post_means, ci,index_n )

post_beta<- as.data.frame(combined[grep('beta',names(post_means)),]) %>%
  filter(index_n<=101)

post_delta<- as.data.frame(combined[grep('delta',names(post_means)),]) %>%
  filter(index_n<=101)

post_epsilon<- as.data.frame(combined[grep('epsilon',names(post_means)),]) %>%
  filter(index_n<=101)

taus<- as.data.frame(combined[grep('tau',names(post_means)),]) 
taus

int<- as.data.frame(combined[grep('int',names(post_means)),]) 
int


post_baseline_prob<- as.data.frame(combined[grep('baseline_prob',names(post_means)),]) 

# ggplot(post_beta, aes(x=index_n, y=exp(post_means))) +
#   theme_classic() +
#   geom_ribbon(aes(x=index_n, ymin=exp(`1`), ymax=exp(`2`)),fill='gray80') +
#   geom_line()+
#   ggtitle('Mortality ratio by Age') +
#     xlab('Age (years)')

p0.age.rate <- ggplot(post_baseline_prob, aes(x=index_n, y=(post_means))) +
  theme_classic() +
  geom_ribbon(aes(x=index_n, ymin=(`1`), ymax=(`2`)),fill='gray80') +
  geom_line()+
    ylab('log-rate of deaths') +
  ggtitle('Log-Deaths/100000 per quarter') +
    xlab('Age (years)')

p0.age.rate

p1.age.jags <- ggplot(post_beta, aes(x=index_n, y=(post_means))) +
  theme_classic() +
  geom_ribbon(aes(x=index_n, ymin=(`1`), ymax=(`2`)),fill='gray80') +
  geom_line()+
    ylab('RR by age') +
  ggtitle('age effect') +
    xlab('Age (years)')

p1.age.jags

# ggplot(post_baseline_prob, aes(x=index_n, y=exp(post_means))) +
#   theme_classic() +
#   geom_ribbon(aes(x=index_n, ymin=exp(`1`), ymax=exp(`2`)),fill='gray80') +
#   geom_line()+
#   ggtitle('Pre-pandemic rate of death') +
#     xlab('Age (years)') +
#   ylab('Deaths/100000 per quarter')

p2.time.age.jags <- ggplot(post_epsilon, aes(x=index_n, y=exp(post_means))) +
  theme_classic() +
  geom_ribbon(aes(x=index_n, ymin=exp(`1`), ymax=exp(`2`)),fill='gray80') +
  geom_line()+
  ggtitle('Time trend by Age') +
    xlab('Age (years)') +
  geom_hline(yintercept=1)+
  ylab('RR per quarter')
p2.time.age.jags

p3.pandemic.age.jags <-  ggplot(post_delta, aes(x=index_n, y=exp(post_means))) +
  theme_classic() +
  geom_ribbon(aes(x=index_n, ymin=exp(`1`), ymax=exp(`2`)),fill='gray80') +
  geom_line()+
  ggtitle('Pandemic effect (RR) by Age') +
    xlab('Age (years)') +
  ylab('Pandemic effect (RR)')+
    geom_hline(yintercept=1)

p3.pandemic.age.jags
```


```{r}
post_beta_samps <-  t(posterior_samples.all[,grep('delta',colnames(posterior_samples.all))])

matplot(post_beta_samps[,1:1000], type='l', col=rgb(0,0,0, alpha=0.1))
```
#3 Direct comparison of INLA and JAGS
```{r}


(p1.age.jags+ggtitle('JAGS') + ylim(0,10)) + (p1.age.inla+ggtitle('INLA')+ ylim(0,10))

#NOTE DIFFERENT SCALE HERE
(p2.time.age.jags+ggtitle('JAGS') +ylim(0.98,1.03)) + (p2.time.age.inla+ggtitle('INLA')+ylim(0.98,1.03))

(p3.pandemic.age.jags+ggtitle('JAGS') +ylim(0.8,1.5)) + (p3.pandemic.age.inla+ggtitle('INLA')+ylim(0.8,1.5))

```

excess deaths and LE

full posterior estimates with uncertainty
```{r, fig.width=8, fig.height=3, eval=F}

post_rate_death <- exp(baseline_prob_post)/100000 #quarterly rate

delta_post <- posterior_samples.all[,grep('delta',colnames(posterior_samples.all))] %>%
  reshape2::melt() %>%
  rename(rep=Var1, delta=value) %>%
  mutate(agey= gsub('delta[', '', Var2, fixed=T),
         agey= as.numeric(gsub(']', '', agey, fixed=T))-1 ,
         pandemic_rr=exp(delta))

life_table_post <- posterior_samples.all[,grep('baseline_prob',colnames(posterior_samples.all))]  %>%
  reshape2::melt() %>%
  rename(rep=Var1, baseline_prob_post=value) %>%
  mutate(agey= gsub('baseline_prob[', '', Var2, fixed=T),
         agey= as.numeric(gsub(']', '', agey, fixed=T))-1 ,
         post_rate_death = exp(baseline_prob_post)/100000,
         annual_prob_death_cond_post = 1- exp(-1*post_rate_death*3), #annual probability of death--meaured 3 quarters of the pandemic https://www.ssa.gov/oact/STATS/table4c6.html
        annual_prob_death_cond_post = if_else(agey==100,1,annual_prob_death_cond_post)) %>% #by kissler's assumption, if survive to 100, probability of death=1
        group_by(rep) %>%
         mutate(prob_death_uncond= calc_pdeath_inst(annual_prob_death_cond_post)) %>%
  ungroup() %>%
  left_join(delta_post, by=c('agey','rep')) %>%
  mutate(rate_death_cond = -1/1*log(1-annual_prob_death_cond_post),
         rate_death_cond_pandemic=rate_death_cond*pandemic_rr ,
         prob_death_cond_pandemic = 1 - exp(-1*rate_death_cond_pandemic),
         excess_cond_prob_death= prob_death_cond_pandemic - annual_prob_death_cond_post,
         
         #Then calculate excess unconditional probability of death
         log_prob_surv_pandemic =log(1-prob_death_cond_pandemic), 
         cum_log_prob_surv_pandemic=cumsum(log_prob_surv_pandemic) - log_prob_surv_pandemic , #log probability survive to this point
         cum_prob_surv_pandemic=exp(cum_log_prob_surv_pandemic), #probability survive to this point
         prob_death_uncond_pandemic= prob_death_cond_pandemic*cum_prob_surv_pandemic,
         prob_death_uncond_pandemic= prob_death_uncond_pandemic/sum(prob_death_uncond_pandemic)
         
         ) #excess probability of death, by age


life_expect_birth_post <-  life_table_post %>%
  mutate(piece = (agey+0.5)*prob_death_uncond) %>%
  group_by(rep) %>%
  summarize(le = sum(piece)) %>%
  ungroup() %>%
  summarize(le =quantile(le, probs=c(0.5, 0.025, 0.975)))
life_expect_birth_post

#function to get cases from m matrix
  rev.m.matrix <- function(mat1) {
    expected <-  reshape2::melt(mat1) %>%
    rename('initial_age'=Var1, index=Var2, uncond_prob=value) %>%
    filter(index >= initial_age) %>%
    arrange(initial_age, index) %>%
    group_by(initial_age) %>%
    mutate(yearN=(row_number() -1),
           age_death= initial_age + yearN  ) %>%
     left_join(pop, by=c("initial_age"="agey") ) 
    return(expected)
  }
  
life_table_post.spl <- split(life_table_post, life_table_post$rep)

expected_deaths_post_fun <- function(ds, k.set=1.2){
  m <- make_upper_tri(ds$prob_death_uncond, shift=FALSE)
  m <- norm_rows(m)
  m_init <- m
  
  pop <- b1 %>% ungroup() %>%
    filter(year==2019 ) %>% select(pop, agey,N_deaths) %>%
    group_by(agey) %>%
    summarize(pop=mean(pop), Obs_deaths2019=sum(N_deaths)) %>%
    ungroup()
  
  #reverse M matrix to get deaths for current and future years
  
  
  m.long <- rev.m.matrix(m_init)  %>%
    mutate("N_deaths_expected" = uncond_prob * pop )
    
  # m.long %>% 
  #   filter(age_death>=65) %>%
  # ggplot( aes(x=yearN, y=N_deaths_expected, group=age_death, color=age_death)) +
  #   geom_line()+
  #   theme_classic()
  
  
  
  #MODIFIED m matrix for pandemic year
  
   k=k.set
  
   # --- Calculate excess probability of death from COVID:
        # Turn the baseline probability of death...
        p_death <- diag(m)
        # into a baseline rate of death...
        r_death <- -log(1-p_death)
        # then calculate the rate of death from COVID...
        xi = life_table$pandemic_rr
        r_death_covid <- xi*r_death
        # and convert back into a probability of death during COVID.
        p_death_covid <- 1-exp(-r_death_covid)
        # Calculate the excess probability of death from COVID
        excess_p_death_covid <- p_death_covid - p_death
        
        # --- Distribute excess probability across future-years: 
        # Create a matrix of future-year proportions: 
        fymat <- m-diag(diag(m))
        # Create a matrix of death-probability dropoff: 
        kmat <- make_upper_tri(c(0,k^(-(0:(ncol(m)-2)))), shift=TRUE)
        # Calculate the normalizing factor for death dropoff: 
        x <- excess_p_death_covid/rowSums(kmat*fymat)
        x[is.na(x)] <- 0
        # Turn this normalizing factor into a matrix:
        xmat <- matrix(rep(x,nrow(m)),nrow=nrow(m))
        # Calculate the distribution of excess death across future-years:
        fy_death_dist_mat <- xmat*kmat*fymat
        # Reduce future-year proportions to account for excess death:
        m_pandemic <- m - fy_death_dist_mat
        # Add that excess death to this year's deaths:
        diag(m_pandemic) <- diag(m_pandemic) + excess_p_death_covid
  
  m.long.pandemic <- rev.m.matrix(m_pandemic)  %>%
    mutate(N_deaths_expected_Post_pand = uncond_prob * pop,
               agec =if_else(age_death>=0 & age_death<18,1, 
                          if_else(age_death>=18 & age_death<40,2,
                          if_else(age_death>=40 & age_death<65,3,
                          if_else(age_death>=65 & age_death<85,4,
                          if_else(age_death>=85 & age_death<101,5, 999)))))) %>%
    ungroup() %>%
    select(agec,N_deaths_expected_Post_pand, age_death, yearN) %>%
    left_join(m.long, by=c('age_death','yearN')) %>%
    group_by(agec, yearN) %>%
    summarize(N_deaths_expected_Post_pand=sum(N_deaths_expected_Post_pand), N_deaths_expected=sum(N_deaths_expected))
  return(m.long.pandemic)
}

expected_deaths_post <- pbapply::pblapply(life_table_post.spl, expected_deaths_post_fun  )


expected_deaths_comb <- bind_rows(expected_deaths_post) %>% 
  filter( yearN<=10 & agec %in% c(2,3,4,5)) %>%
  group_by(yearN, agec) %>%
  summarize(N_deaths_expected_Post_pand = quantile(N_deaths_expected_Post_pand, probs=c(0.5)),
            N_deaths_expected_Post_pand_lcl = quantile(N_deaths_expected_Post_pand, probs=c(0.025)),
            N_deaths_expected_Post_pand_ucl = quantile(N_deaths_expected_Post_pand, probs=c(0.975)),
            N_deaths_expected = quantile(N_deaths_expected, probs=c(0.5)),
            N_deaths_expected_lcl = quantile(N_deaths_expected, probs=c(0.025)),
            N_deaths_expected_ucl = quantile(N_deaths_expected, probs=c(0.975)),
            )
  
saveRDS(expected_deaths_comb,'./Results/expected_deaths_comb.rds')
```

```{r}
expected_deaths_comb <- readRDS('./Results/expected_deaths_comb.rds')

ggplot(expected_deaths_comb, aes(x=yearN, y=N_deaths_expected_Post_pand) ) +
  geom_line(col='red')+
  geom_line(aes(x=yearN, y=N_deaths_expected), col='gray',lty=2) +
  theme_classic() +
  ylim(0,NA) +
  facet_wrap(~agec, scales='free', nrow=1)


```


```{r}
##########################

###DO NOT MODIFY--this is run with just mean; not full posterior samples

rate_death <- exp(post_baseline_prob$post_means)/100000 #quarterly rate


#conditional probability of death per year
annual_prob_death <- 1- exp(-1*rate_death*3) #annual probability of death; compare to: https://www.ssa.gov/oact/STATS/table4c6.html; measre 3 quarters
annual_prob_death[101] <- 1 #by kissler's assumption, if survive to 100, probability of death=1

post_delta_rr <- post_delta %>%
    mutate(agey= index_n -1) %>%
  mutate(pandemic_rr=exp(post_means),pandemic_rr_lcl=exp(`1`),pandemic_rr_ucl=exp(`2`) ) %>%
  select(agey,starts_with('pandemic_rr'))

life_table <- cbind.data.frame('prob_death_cond'=annual_prob_death) %>% #prob of dying in this year conditional on having suvived
  mutate(agey=0:100,
         prob_death_uncond= calc_pdeath_inst(prob_death_cond)
         
        ) %>% #unconditional probability of dying at this age
  left_join(post_delta_rr, by='agey') %>%
  mutate(rate_death_cond = -1/1*log(1-prob_death_cond),
         rate_death_cond_pandemic=rate_death_cond*pandemic_rr ,
         prob_death_cond_pandemic = 1 - exp(-1*rate_death_cond_pandemic),
         excess_cond_prob_death= prob_death_cond_pandemic - prob_death_cond,
         
         #Then calculate excess unconditional probability of death
         log_prob_surv_pandemic =log(1-prob_death_cond_pandemic), 
         cum_log_prob_surv_pandemic=cumsum(log_prob_surv_pandemic) - log_prob_surv_pandemic , #log probability survive to this point
         cum_prob_surv_pandemic=exp(cum_log_prob_surv_pandemic), #probability survive to this point
         prob_death_uncond_pandemic= prob_death_cond_pandemic*cum_prob_surv_pandemic,
         prob_death_uncond_pandemic= prob_death_uncond_pandemic/sum(prob_death_uncond_pandemic,na.rm=T)
         
         ) #excess probability of death, by age


life_expect_birth <-  life_table %>%
  mutate(piece = (agey+0.5)*prob_death_uncond) %>%
  summarize(le = sum(piece))

plot(life_table$prob_death_cond, type='l' ,bty='l')
points(life_table$prob_death_cond_pandemic, type='l', col='red', lty=2)

#Unconditional probability of dying by age; by definition, if shift earlier, fewer people will die later
p1 <-life_table %>%
  filter(agey<100) %>%
 ggplot( aes(x=agey, y=prob_death_uncond)) +
  theme_classic()+
  geom_line(col='gray') +
  geom_line(aes(x=agey, y=prob_death_uncond_pandemic), col='red')
p1

p2 <-life_table %>%
  filter(agey<100) %>%
 ggplot( aes(x=agey, y=prob_death_cond)) +
  theme_classic()+
  geom_line(col='gray') +
  geom_line(aes(x=agey, y=prob_death_cond_pandemic), col='red')
p2



plot(life_table$cum_prob_surv, type='l')
points(life_table$cum_prob_surv_pandemic, type='l', col='red', lty=2)


# Create a life matrix using the baseline probabilities of death: 
m <- make_upper_tri(life_table$prob_death_uncond, shift=FALSE)
m <- norm_rows(m)
m_init <- m

pop <- b1 %>% ungroup() %>%
  filter(year==2019 ) %>% select(pop, agey,N_deaths) %>%
  group_by(agey) %>%
  summarize(pop=mean(pop), Obs_deaths2019=sum(N_deaths)) %>%
  ungroup()

rev.m.matrix <- function(mat1) {
  expected <-  reshape2::melt(mat1) %>%
  rename(index=Var2, uncond_prob=value) %>%
  mutate(initial_age=Var1-1) %>%
  filter(index >= (initial_age+1) ) %>%
  arrange(initial_age, index) %>%
  group_by(initial_age) %>%
  mutate(yearN=(row_number() -1),
         age_death= initial_age + yearN ) %>%
    filter(age_death<=100) %>%
   left_join(pop, by=c("initial_age"="agey") ) %>% #match pop to initial age of cohort
  
  return(expected)
}

#reverse M matrix to get deaths for current and future years


m.long <- rev.m.matrix(m_init)  %>%
  mutate("N_deaths_expected" = uncond_prob * pop )
  
m.long %>% filter(yearN==0) %>%
  ggplot(aes(x=age_death, y=Obs_deaths2019))+
  geom_line() +
  geom_line(aes(x=age_death, y=N_deaths_expected*4/3,), col='red')


m.long %>% 
  filter(age_death>=65) %>%
ggplot( aes(x=yearN, y=N_deaths_expected, group=age_death, color=age_death)) +
  geom_line()+
  theme_classic()



#MODIFIED m matrix for pandemic year

 k=0.1

 # --- Calculate excess probability of death from COVID:
      # Turn the baseline probability of death...
      p_death <- diag(m)
      # into a baseline rate of death...
      r_death <- -log(1-p_death)
      # then calculate the rate of death from COVID...
      xi = life_table$pandemic_rr
      r_death_covid <- xi*r_death
      # and convert back into a probability of death during COVID.
      p_death_covid <- 1-exp(-r_death_covid)
      # Calculate the excess probability of death from COVID
      excess_p_death_covid <- p_death_covid - p_death
      
      # --- Distribute excess probability across future-years: 
      # Create a matrix of future-year proportions: 
      fymat <- m-diag(diag(m))
      
      # Create a matrix of death-probability dropoff: 
      kmat <- make_upper_tri(c(0,k^(-(0:(ncol(m)-2)))), shift=TRUE)
      
      # Calculate the normalizing factor for death dropoff: 
      x <- excess_p_death_covid/rowSums(kmat*fymat)
      x[is.na(x)] <- 0
      # Turn this normalizing factor into a matrix:
      xmat <- matrix(rep(x,nrow(m)),nrow=nrow(m))
      # Calculate the distribution of excess death across future-years:
      fy_death_dist_mat <- xmat*kmat*fymat
      # Reduce future-year proportions to account for excess death:
      m_pandemic <- m - fy_death_dist_mat
      # Add that excess death to this year's deaths:
      diag(m_pandemic) <- diag(m_pandemic) + excess_p_death_covid

m.long.pandemic <- rev.m.matrix(m_pandemic)  %>%
  mutate(N_deaths_expected_Post_pand = uncond_prob * pop,
             agec =if_else(age_death>=0 & age_death<18,1, 
                        if_else(age_death>=18 & age_death<40,2,
                        if_else(age_death>=40 & age_death<65,3,
                        if_else(age_death>=65 & age_death<85,4,
                        if_else(age_death>=85 & age_death<=100,5, 999)))))) %>%
  ungroup() %>%
  select(agec,N_deaths_expected_Post_pand, age_death, yearN) %>%
  left_join(m.long, by=c('age_death','yearN')) %>%
  group_by(agec, yearN) %>%
  summarize(N_deaths_expected_Post_pand=sum(N_deaths_expected_Post_pand, na.rm=T), N_deaths_expected=sum(N_deaths_expected, na.rm=T))

#Plot
p1 <- m.long.pandemic %>% 
  filter( yearN<=10 & agec %in% c(2,3,4,5)) %>%
ggplot( aes(x=yearN, y=N_deaths_expected_Post_pand) ) +
  geom_line(col='red')+
  geom_line(aes(x=yearN, y=N_deaths_expected), col='gray',lty=2) +
  theme_classic() +
  ylim(0,NA) +
  facet_wrap(~agec, scales='free', nrow=1)

p1


```


## Competing risks

People can die from COVID, or they can die from something else. Probability of death from something else can be determined from baseline death rate

start with N people, each quarter, probability of being alive is (1-P_die) for 3 quarters it (1-P_die)^3 

** important note: as in Kissler and grad,ned to convert probabilities to rates before multiplying with RR, then convert back to probability ** 

```{r}

P_die_qtr_base <- exp(7.5302187)/100000 # probability an 84 year old dies in a quarter

rate_die_qtr_base <- -log(1-P_die_qtr_base) #convert probability to rate
  
RR_Pandemic <- exp(0.164847030)

rate_die_qtr_pandemic <- rate_die_qtr_base*RR_Pandemic #multiple rate by RR

P_die_qtr_pandemic <- 1 - exp(-rate_die_qtr_pandemic*1) #convert rate back to probability

P_die_virus <- P_die_qtr_pandemic - P_die_qtr_base #Prob excess deaths; most of which are virus

P_die_year <- 1 - (1 - P_die_qtr_base)^4 #0.072 pretty close to government estimate of 0.064 for women and 0.085 for men https://www.ssa.gov/oact/STATS/table4c6.html#fn1

P_die_pandemic <- 1 - (1 - P_die_qtr_pandemic)^4


#Category 1= alive

#Category 2= die covid

#Category 3= die other

#Start with a cohort of 1000

t <- 1:40 #  10 years of quarterly data

Alive <- rep(NA, length(t))
Die_COVID <- rep(NA, length(t))
Die_Other <- rep(NA, length(t))

Alive[1] <- 1000
Die_COVID[1] <- 0
Die_Other[1] <- 0

for(i in 2:length(t)) {
  
  Alive[i] <- Alive[i-1] - Alive[i-1] * P_die_qtr_base  - Alive[i-1] * P_die_virus
  Die_COVID[i] <- Die_COVID[i-1] + Alive[i-1] * P_die_virus #cumulative deaths covid
  Die_Other[i] <- Die_Other[i-1] + Alive[i-1] * P_die_qtr_base #cumulative deaths other

}

plot(Die_COVID)

```
 
 It might make sense to focus in on people with certain conditions, like cancer; diabetes; etc.
 Need prevalence info by age and ethnicity; could get from the VA
 
```{r}
#should be 282,801 diabetes deaths in US in 2019 according to CDC wonderi in E10-E14
d1 <- readRDS('./Data/diabetes_covid_deaths_line_list_no_geo.rds')

 d1 %>%
  filter(year %in% c(2019,2020)) %>%
   dplyr::group_by(year) %>%
  dplyr::summarize(N_diabetes=sum(diabetes), N_covid_diabetes=sum(covid_diabetes))

 
summary1 <- d1 %>%
  group_by(age_group,race_recode,year) %>%
  filter(year %in% c(2018,2019,2020)) %>%
  summarize(diabetes=sum(diabetes), covid_diabetes=sum(covid_diabetes)) %>%
  reshape2::dcast( race_recode +age_group ~ year, value.var='diabetes') %>%
  mutate(ratio20_19=`2020`/`2019`, ratio19_18=`2019`/`2018`)
  
#This plots shows little year-over-year increase 2018-2019, larges in crease 2019-2020 (not just secular trend)
plot(summary1$ratio20_19, summary1$ratio19_18)
#View(summary1)

```
 
 